---
title: "PBMExperiment"
author: Patrick Kimes
output:
  github_document:
    toc: true
    toc_depth: 2
    fig_width: 12
    fig_height: 4
    dev: png
---

```{r echo = FALSE}
gpr_dir <- ".../gpr_files"
```

**THIS PACKAGE IS STILL UNDER ACTIVE DEVELOPMENT**

# Introduction

This package provides tools for reading, plotting, and analyzing protein binding microarray (PBM) data in *R*. Currently, input data must be available as GenePix Results (GPR) files. A more stable and separate set of tools developed by Berger and Bulyk (Nature Protocols, 2009) for the analysis of PBMs is available for download at the [_Universal PBM Analysis Suite_ site](the_brain.bwh.harvard.edu/PBMAnalysisSuite/indexSep2017.html).

Several functions for loading and preprocessing microarray data are available in *R* through the widely used [*affy*]() and [*limma*]() packages in Bioconductor. Additionally, the *limma* package also includes several functions for testing differential expression with gene expression arrays. Rather than reinventing the (entire) wheel, this package (*PBMExperiment*) includes several easy-to-use wrapper functions to the preprocessing functions implemented in both *affy* and *limma* (prefixed `rma*` and `limma*`, respectively). Due to the specialized nature of PBMs, the *PBMExperiment* package includes specialized procedures for probe-level quantification not implemented in either of the aforementioned packages. 

# Installation

To install the latest version of this package from GitHub, use the following command from the `devtools` package:

```{r, eval = FALSE}
devtools::install_github("pkimes/PBMExperiment")
```

If the package is already installed, there is no need to re-install the package except to check for updates.
After installing, the package still needs to be loaded to make the functions available in R.

```{r}
library("PBMExperiment")
```

# Quickstart

Suppose we have a folder, `gpr_dir`, containing GPR files for a single experiment. We first determine the set of samples in that folder with `buildPBMTable`, and then read in the samples using `buildPBMExperiment`.

```{r}
samples <- buildPBMTable(gpr_dir)
pbmdata <- buildPBMExperiment(samples, probes = pbm_8x60k_v1)
```

The data is now available for plotting and analysis.
Here, we show an example of computing the spatial bias in each sample, and plotting the bias across the grid of probe coordinates.

```{r}
pbm_adjusted <- spatiallyAdjust(pbmdata)
```

```{r, fig.width = 12, fig.height = 8}
pbmPlotChip(pbm_adjusted, assay_name = "spatialbias",
            log_scale = FALSE, .facet = FALSE) +
    facet_grid(lp ~ condition)
```

For more details on the available plotting and analysis functions, read the more detailed sections below.

# Overview

## Reading GPR Data

Again, we start by identifying the set of sample GPRs in the directory, `gpr_dir`. In the quickstart example above, only raw `"Alexa488"` GPR scans were captured (by default). However, there may also be Cy3 and Masliner GPR files in the same directory. To also capture these samples, we can change the `gpr_type=` parameter to any of `"Masliner"`, `"Cy3"`, or `"all"`.

```{r}
samples_all <- buildPBMTable(gpr_dir, gpr_type = "all")
```

The GPR scan type is included in the `scan` column of the sample table.

```{r}
samples_all$scan
```

Given the table of samples, we next load the GPR data into R according to the `gpr` column of the sample table. Depending on the number of samples, this may take a few seconds to a few minutes. For illustration, we only load a subset of the samples.

```{r}
condition_set <- c("HOXD13-REF", "HOXD13-Q325R", "HOXD13-I322L")
pbmdata <- buildPBMExperiment(samples_all[samples_all$lp %in% c(500, 550) &
                                          samples_all$scan == "Alexa488" &
                                          samples_all$condition %in% condition_set, ],
                              probes = pbm_8x60k_v1)
```

The sample data is stored as a [_SummarizedExperiment_](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html), a common structure for storing genomic data in [Bioconductor](http://bioconductor.org/).

```{r}
pbmdata
```

A _SummarizedExperiment_ is made up of three parts: **1.** assays (primary data), **2.** row data, and **3.** column data.  

Naturally, the probe-level intensities from the GPRs are available in the `"gpr"` **assay** as a DataFrame with rows and columns corresponding to probes and samples.

```{r}
assay(pbmdata, "gpr")[1:5, 1:5]
```

The **row data** contains information about the rows (probes).

```{r}
rowData(pbmdata)[1:5, ]
```

Similarly, the **column data** contains information about the columns (samples).

```{r}
colData(pbmdata)[1:5, ]
```

Notice that when we loaded the GPR data, we also specified the probe design used in this experiment, `pbm_8x60k_v1`. This default probe set is included with the package. The probe information is matched with the GPR data according to Row and Column coordinates.

```{r}
head(pbm_8x60k_v1)
```

Notice that this information is included in the row data shown above.  

## Plotting GPR Data

Now that we have the data, we can use plots to visually explore the distributions of probe intensities across samples. Currently, the following functions are available for plotting GPR data: `pbmPlotDensity()`, `pbmPlotChip()`, `pbmPlotScatter()`, and `pbmPlotComparison()`. Each function returns a `ggplot2` object that can be customized to generate more specialized plots. By default, each function plots each sample separately ("facets") according to the `condition` column of the `colData`.

```{r}
colData(pbmdata)$condition
```

If `condition` is not unique to each sample, as is the case here, we can set `.facet = FALSE` in the function, and explicitly facet the returned plot. Faceting can be performed using any of the columns in the `colData` of the SummarizedExperiment.

```{r}
colData(pbmdata)
```

In this example, laser power (`lp`) and `condition` uniquely identify each sample, and can be used for faceting the samples.  

First, we use `lp` and `condition` to facet a plot of the distribution of probe intensities for each sample. 

```{r, fig.width = 10, fig.height = 6}
pbmPlotDensity(pbmdata, .facet = FALSE) +
    facet_grid(lp ~ condition)
```

Similarly, we can plot probe intensities for each sample across the coordinates of the array. 

```{r, fig.width = 10, fig.height = 6}
pbmPlotDensity(pbmdata, .facet = FALSE) +
    facet_grid(lp ~ condition)
```

Since these experiments typically include a reference sample, another useful view of the data is to plot probe intensities across conditions against a reference sample. This may be useful to, for example, compare across conditions for a fixed laser power (`lp`) or across laser powers for a single condition (`condition`). Since the ultimate goal is to compare across samples, we plot the former. To do this, we simply subset the samples (columns) according to the `colData`, and call the plotting function. 

```{r, fig.width = 10, fig.height = 6}
pbmPlotScatter(pbmdata[, colData(pbmdata)$lp == 500])
```

A similar but alternative view of the data is the [MA plot](https://en.wikipedia.org/wiki/MA_plot), which plots the mean log-intensity against the difference in log-intensity for each pair of samples.

```{r, fig.width = 15, fig.height = 5}
pbmPlotScatter(pbmdata[, colData(pbmdata)$lp == 500], maplot = TRUE)
```

Finally, often, we may be interested in comparing across replicates of the same experiment. Here, we illustrate the functionality by comparing conditions across two laser powers. Each sample is matched using a column in the `colData` specified by `match_by = `. Note that the input to `match_by = ` should *not* be quoted.

```{r, fig.width = 15, fig.height = 5}
pbmPlotComparison(pbmdata[, colData(pbmdata)$lp == 500],
                  pbmdata[, colData(pbmdata)$lp == 550],
                  match_by = condition)
```

As above, MA plots are also implemented.

```{r, fig.width = 15, fig.height = 5}
pbmPlotComparison(pbmdata[, colData(pbmdata)$lp == 500],
                  pbmdata[, colData(pbmdata)$lp == 550],
                  match_by = condition,
                  maplot = TRUE)
```

## Pre-Processing GPR Data

**Choice of pre-processing steps for PBM data is still under investigation.**

Approaches for pre-processing of microarray intensities has been widely studied over the past 15 years. Several popular and successful approaches, including RMA, have been implemented and made available through the *affy* and *limma* packages. However, the functions implemented within these packages expect slightly different inputs and generate slighlty different outputs, making it difficult to directly apply the methods to our data stored in *SummarizedExperiment* class. To make access to these functions easier, we have written several wrapper functions to pre-processing functions implemented in both packages.

Single-channal pre-processing includes two primary steps: **within-sample background correction** and **between-sample normalization**. The corresponding functions implemented in *limma* are wrapped as `limmaBackgroundCorrection` and `limmaNormalizeBetweenSamples`. Each function returns the same *SummarizedExperiment* object, but with normalized expression estimates.

```{r}
pbmdata_bc <- limmaBackgroundCorrect(pbmdata)
```

Since the returned object is again a SummarizedExperiment, we can easily plot the data using the functions described above. Here, we compare the background corrected and original data for scans at laser power 500.

```{r, fig.width = 15, fig.height = 5}
pbmPlotComparison(pbmdata[, colData(pbmdata)$lp == 500],
                  pbmdata_bc[, colData(pbmdata_bc)$lp == 500])
```

For provenance, the correction step is also noted in the `metadata` of the object.

```{r}
metadata(pbmdata_bc)
```

We next apply cross-sample normalization.

```{r}
pbmdata500_norm <- limmaNormalizeBetweenArrays(pbmdata_bc[, colData(pbmdata)$lp == 500])
```

Again, we compare the normalized data against the original data.

```{r, fig.width = 15, fig.height = 5}
pbmPlotComparison(pbmdata_bc[, colData(pbmdata)$lp == 500],
                  pbmdata500_norm,
                  maplot = TRUE)
```

Next, we are ready to convert the probe-level intensities to candidate motif-level intensity estimates. 

## Quantifying GPR Data



# References

- [Berger MF and Bulyk ML. "Universal protein-binding microarrays for the comprehensive characterization of the DNA-binding specificities of transcription factors." Nature Protocols. (2009) 4(3):393-411.](http://the_brain.bwh.harvard.edu/pubs/uPBMs_NatProt09.pdf)
- [Berger MF, Philippakis AA, Qureshi A, He FS, Estep PW III and Bulyk ML. "Compact, universal DNA microarrays to comprehensively determine transcription-factor binding site specificities." Nature Biotechnology. (2006) 24(11):1429-1435.](http://the_brain.bwh.harvard.edu/pubs/CompactDNAview.pdf)
- [Universal PBM Analysis Suite](the_brain.bwh.harvard.edu/PBMAnalysisSuite/indexSep2017.html)
- [Morgan M, Obenchain V, Hester J and Pagès H (2017). SummarizedExperiment: SummarizedExperiment container. R package version 1.8.1.](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html)
