---
title: "upbm: Inference Details"
author: "Patrick K. Kimes"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('upbm')`"
output:
  BiocStyle::html_document:
    highlight: pygments
    toc: true
    fig_width: 12
    fig_height: 4
bibliography: library.bib
vignette: >
  %\VignetteIndexEntry{5. upbm: inference}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r echo=FALSE, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE, cache = TRUE, dev = "png",
                      message = FALSE, error = FALSE, warning = TRUE)
```

# Introduction

For an introduction to the *upbm* package, please see the quick start vignette (`vignette("upbm-quickstart")`). Here, we provide details on the various testing procedures briefly referenced in the quick start vignette. We assume basic knowledge of the _PBMExperiment_ class (see `vignette("upbm-classes")`). The same example HOXC9 dataset from the *upbmData* package is also used in this vignette to illustrate the various functions.

The _upbm_ package includes three testing functions which all take the output of `upbm::kmerFit` and return a table of 8-mer-level effect sizes and significance estimates. In the quick start vignette, only a single testing procedure was covered, namely, differential affinity testing using `upbm::kmerTestContrast`. The three testing procedures are:

- preferential affinity testing (`upbm::kmerTestAffinity`),
- differential affinity testing (`upbm::kmerTestContrast`),
- differential specificity testing (`upbm::kmerTestSpecificity`).

We cover each procedure in detail in this vignette.

```{r}
suppressPackageStartupMessages(library("upbm"))
```

Throughout, we will be using `broom::tidy` to tidy data stored in *PBMExperiment* and *SummarizedExperiment* objects for interactive analysis and plotting (see `vignette("other-tidydata")`). We load the *dplyr* package to help with interactive analysis.

```{r}
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("ggplot2"))
```

# HOXC9 Dataset

For details on the example HOXC9 dataset, see the quick start vignette in this package or the *upbmData* package vignette. Here, we will use both Alexa488 and Cy3 scans. 

```{r load-hoxc9}
data(hoxc9alexa, package = "upbmData")
data(hoxc9cy3, package = "upbmData")
```

Here, we again use just single PMT gain scan for the Alexa488 scans.

```{r subset-pmt}
alexa_subset <- hoxc9alexa[, colData(hoxc9alexa)$pmt == 450]
```

We also load the reference Cy3 dataset.

```{r load-cy3ref}
data(refcy3_8x60k_v1, package = "upbmAux")
```

# Sample Processing 

We will perform the earlier steps of pre-processing using the `upbm::pmbPreprocess` wrapper. Details on pre-processing steps are covered in an earlier vignette (see `vignette("analysis-preprocessing"`).

```{r preprocess}
alexa_subset <- upbmPreprocess(pe = alexa_subset,
                               cy3pe = hoxc9cy3,
                               cy3refpe = refcy3_8x60k_v1,
                               verbose = FALSE)
```

We will further summarize results to 8-mer-level affinity estimates with `HOXC9-REF` as the baseline condition. As above, details on summarization steps are covered in an earlier vignette (see `vignette("analysis-summarization"`).

```{r summarize}
alexa_pfit <- probeFit(alexa_subset)
alexa_kfit <- kmerFit(alexa_pfit, baseline = "HOXC9-REF")
```

# Preferential Affinity

In contrast to the two other testing procedures which consider **differential** behavior across conditions, preferential affinity considers each condition separately. For each condition, we apply a testing procedure to determine the set of 8-mers _preferentially_ bound by the condition. The `upbm::kmerTestAffinity` function only takes the single _PBMExperiment_ returned by `upbm::kmerFit` as input. 

```{r prefaff}
alexa_pa <- kmerTestAffinity(alexa_kfit)
alexa_pa
```

The returned *PBMExperiment* object contains the results of affinity testing in several assays along with many of the same assays as the input.

```{r prefaff-assays}
assayNames(alexa_pa)
```

Importantly, the new `affinitySignal` assay contains the estimated "signal" component of the K-mer affinity (`affinityEstimate` assay). The signal component is computed by fitting a normal and expontential convolution to the K-mer affinity estimates and taking the expected exponential component. This is similar to a step included in the RMA approach to microarray data normalization [@irizarry2003]. The `affinityZ`, `affinityP` and `affinityQ`, contain the _z_-score, un-adjusted and adjusted _p_-values from testing the null hypothesis that `affinitySignal` is zero. 

The distribution of `affinityEstimates` can be colored by the adjusted _p_-values (`affinityQ`) to show the distribution of preferentially bound 8-mers in each condition. We bin the significance values for cleaner plotting.

```{r prefaff-plot, fig.width = 10, fig.height = 3.5}
pa_dat <- broom::tidy(alexa_pa, c("affinityEstimate", "affinityQ"))

pa_dat %>%
    dplyr::mutate(affinityQ = cut(affinityQ, c(1e0, 1e-2, 1e-3, 1e-6, 0),
                                  include.lowest = TRUE),
                  affinityQ = factor(affinityQ, levels = rev(levels(affinityQ)))) %>%
    ggplot(aes(x = affinityEstimate, fill = affinityQ)) +
    geom_histogram(binwidth = .1, boundary = 10, color = 'black', position = "stack", alpha = 1) +
    scale_fill_brewer("q-value", palette = "Spectral", direction = -1, drop = FALSE,
                      na.value = "black") +
    theme_bw() +
    facet_wrap( ~ cname, nrow = 1, scales = "free_y") +
    ggtitle("HOXC9 8-mer preferential affinity")
```

# Differential Affinity

Differential affinity is likely to most intuitive of the three testing procedures. With differential affinity, we are interested in identifying the set of 8-mers with differential signal across PBM samples between the specified baseline and all other conditions. The underlying hypothesis we test here is whether the estimated affinity differences are zero or not. 

The `upbm::kmerTestContrast`, again, only takes the single _PBMExperiment_ returned by `upbm::kmerFit` as input. 

```{r difaff}
alexa_da <- kmerTestContrast(alexa_kfit)
alexa_da
```

The returned *PBMExperiment* object contains the results of differential affinity testing a few new assays. The `contrastZ`, `contrastP`, and `contrastQ` contain the _z_-score, and unadjusted and adjusted _p_-values for the test at each 8-mer. 

```{r difaff-assays}
assayNames(alexa_da)
```

Here, the _z_-score is simply the ratio of the `contrastDifference` and `contrastVariance` computed during the `upbm::kmerFit` call. We can again plot the results as an MA plot and color 8-mers by significance level. We also remember to exclude the baseline condition, `HOXC9-REF`, from the table as all of the values are `NA`s.

```{r difaff-plot, fig.width = 10, fig.height = 3.5}
da_dat <- broom::tidy(alexa_da, c("contrastAverage", "contrastDifference",
                                  "contrastQ"))

da_dat <- da_dat %>%
    dplyr::filter(cname != "HOXC9-REF") %>%
    dplyr::mutate(contrastQ = cut(contrastQ, c(1e0, 1e-2, 1e-3, 1e-6, 0),
                                  include.lowest = TRUE),
                  contrastQ = factor(contrastQ, levels = rev(levels(contrastQ))))

da_dat %>%
    ggplot(aes(x = contrastAverage, y = contrastDifference,
               color = contrastQ)) +
    geom_point(alpha = 1/10) +
    scale_color_brewer("q-value", palette = "Spectral", direction = -1, drop = FALSE,
                       na.value = "black") +
    geom_hline(color = 'gray', yintercept = 0) + 
    theme_bw() +
    guides(color = guide_legend(override.aes = list(alpha = 1))) +
    facet_grid(. ~ cname) +
    ggtitle("HOXC9 8-mer differential affinity")
```

We can see that few 8-mers are called differential for most of the conditions. More formally, we can count up the number of 8-mers at each of the different significance thresholds above.

```{r difaff-tab}
da_dat %>%
    dplyr::count(cname, contrastQ) %>%
    tidyr::spread(contrastQ, n, fill = 0)
```

# Differential Specificity

Differential specificity, while similar to affinity is treated separately. In some comparisons, the two proteins being contrasted may have completely different binding preferences, preferring different subsets of 8-mers. We call this phenomenon _differential specificity_. 

In the MA plot above, we see that the general trend for all contrasts is a single trajectory, even for the R222W allele. This is the result of shared specificity, even with decreased affinity. In contrast, when differential specificity is present, rather than follow a single trajectory, the MA plot will appear to split, with 8-mers preferential in the two condition spreading. 

Intuitively, testing for differential specificity is performed by first fitting a single trajectory to the MA plot, followed by testing for deviations from the trajectory. Since deviations from the baseline condition are often of more interest, rather than using the average between conditions as the x-axis, the trajectory is fit with the reference affinity along the x-axis. This can be changed by setting `useref = FALSE` in the call to `upbm::kmerTestSpecificity`.

```{r difspec}
alexa_ds <- kmerTestSpecificity(alexa_kfit)
alexa_ds
```

By default, the trajectory is fit using cubic B-splines with 3 knots placed evenly spaced along the range of the x-axis. Alternatively, a LOESS smoother can be fit by specifying `method = "loess"`. 

The returned *PBMExperiment* includes several new assays.

```{r difspec-assays}
assayNames(alexa_ds)
```

The x-axis used for fitting the specificity trend is stored in the `specificityAxis` assay. The fit and residuals are included in the `contrastFit` and `contrastResidual` assays. Finally, the `specificityZ`, ``specificityP`, and `specificityQ` assays contain the _z_-score, unadjusted and adjusted _p_-values from testing the null hypothesis that `contrastResidual` is zero. 

We can plot the data along with the `contrastFit` to see the B-spline trajectories fit to the data.

```{r difspec-plot, fig.width = 10, fig.height = 3.5}
ds_dat <- broom::tidy(alexa_ds, c("specificityAxis", "specificityQ",
                                  "contrastDifference", "contrastFit",
                                  "contrastResidual"))

ds_dat <- ds_dat %>%
    dplyr::filter(cname != "HOXC9-REF") %>%
    dplyr::mutate(specificityQ = cut(specificityQ, c(1e0, 1e-2, 1e-3, 1e-6, 0),
                                  include.lowest = TRUE),
                  specificityQ = factor(specificityQ, levels = rev(levels(specificityQ))))

ds_dat %>%
    ggplot(aes(x = specificityAxis)) +
    geom_point(aes(y = contrastDifference, color = specificityQ), alpha = 1/10) +
    theme_bw() +
    geom_line(aes(y = contrastFit), color = 'black') + 
    scale_color_brewer("q-value", palette = "Spectral", direction = -1, drop = FALSE,
                       na.value = "black") +
    geom_hline(yintercept = 0, color = 'gray') + 
    guides(color = guide_legend(override.aes = list(alpha = 1))) +
    facet_grid(. ~ cname) +
    ggtitle("HOXC9 8-mer differential specificity")
```

We see that unlike differential affinity, very few 8-mers are called in the differential specificity analysis for all alleles.

```{r difspec-tab}
ds_dat %>%
    dplyr::count(cname, specificityQ) %>%
    tidyr::spread(specificityQ, n, fill = 0)
```

We can similarly plot the residuals as well.

```{r difspec-plot2, fig.width = 10, fig.height = 3.5}
ds_dat %>%
    ggplot(aes(x = specificityAxis)) +
    geom_point(aes(y = contrastResidual, color = specificityQ), alpha = 1/10) +
    geom_hline(yintercept = 0, color = 'gray') + 
    theme_bw() +
    scale_color_brewer("q-value", palette = "Spectral", direction = -1, drop = FALSE,
                       na.value = "black") +
    guides(color = guide_legend(override.aes = list(alpha = 1))) +
    facet_grid(. ~ cname) +
    ggtitle("HOXC9 8-mer differential specificity (residuals)")
```

# References
