---
title: "upbm: Class Details"
author: "Patrick K. Kimes"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('upbm')`"
output:
  BiocStyle::html_document:
    highlight: pygments
    toc: true
    fig_width: 12
    fig_height: 4
bibliography: library.bib
vignette: >
  %\VignetteIndexEntry{2. upbm: classes}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r echo=FALSE, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE, cache = TRUE, dev = "png",
                      message = FALSE, error = FALSE, warning = TRUE)
```

# Introduction

For an introduction to the *upbm* package, please see the quick start vignette (`vignette("upbm-quickstart")`). Here, we provide details on the two classes introduced in the package, _PBMDesign_ and _PBMExperiment_. The _PBMDesign_ class acts as a container for protein binding microarray (PBM) design information, while the _PBMExperiment_ class serves as the container for PBM experimental data. 

```{r}
suppressPackageStartupMessages(library("upbm"))
```

_PBMDesign_ objects for standard universial PBM (uPBM) designs are included in the _upbmAux_ package. Additionally, example _PBMExperiment_ objects are included in the _upbmData_ package containing data from uPBM experiments performed using the HOXC9 transcription factor. We use these objects to help illustrate the simple structure of each class.

```{r}
suppressPackageStartupMessages(library("upbmAux"))
suppressPackageStartupMessages(library("upbmData"))
```

# PBMDesign

The _PBMDesign_ class is used to organize the probe information for PBM experiments. _PBMDesign_ objects are composed of 3 components (_slots_):

- `design`: a data.frame of probe IDs, sequences, and other metadata, 
- `probeFilter`: an optional list of filtering rules to distinguish signal probes from control spots,
- `probeTrim`: an optional pair of integers specifying the start and end of the probe sequences to use for analysis.

We can take a look at the `pbm_8x60k_v1` design object from the _upbmAux_ package to get a better sense of these components.

```{r pbmdesign-load-example}
data(pbm_8x60k_v1, package = "upbmAux")
pbm_8x60k_v1
```

From the object description, we can see that the design contains `r nrow(design(pbm_8x60k_v1))` rows and `r ncol(design(pbm_8x60k_v1))` columns. Each row is a probe in the PBM array. Two columns, `"probeID"` and `"Sequence"` **must** be defined for the object to be a valid _PBMDesign_. The other two columns, `"Row"` and `"Column"` are optional information which are needed to perform spatial adjustment with `upbm::spatiallyAdjust`. 

```{r pbmdesign-show-design}
head(design(pbm_8x60k_v1))
```

In addition to dsDNA probes which measure protein binding, PBMs typically include control probes designed by Agilent or the user. These probes should be excluded for certain normalization and downstream analysis steps (e.g. Cy3 normalization and spatial adjustment). 

The `probeFilter` slot is a list of functions which are used to distinguish target probes from control probes. The list must be named with names matching columns in the design shown above. We can get and set the `probeFilter` slot by calling `upbm::probeFilter`.

```{r pbmdesign-show-probefilter-names}
names(probeFilter(pbm_8x60k_v1))
```

Here, the _PBMDesign_ object includes a single probe filtering rule associated with the `"probeID"` column. 

```{r pbmdesign-show-probefilter-rule}
probeFilter(pbm_8x60k_v1)$probeID
```

The rule must be a function that takes the specified column (`"probeID"`) as input and returns a logical vector of the same length. Only rows which return `TRUE` for all rules are kept for downstream analysis. In the `pbm_8x60k_v1` design, the single filter checks the probe IDs and returns `TRUE` for the subset prefixed with `dBr_`, corresponding to the subset of de Bruijn sequence probes on the array. Filtering can be performed using `upbm::pbmFilterProbes`.

```{r pbmdesign-pbmfilterprobes}
pbm_8x60k_filtered <- pbmFilterProbes(pbm_8x60k_v1)
pbm_8x60k_filtered
```

Notice that the returned _PBMDesign_ has fewer probes than the original design. We can explicitly verify that this smaller number of probes matches the number of probes containing the `dBr` prefix in the `"probeID"` column.

```{r pbmdesign-checkfilter}
table(grepl("^dBr", design(pbm_8x60k_v1)$probeID))
```

After filtering has been performed, the `probeFilter` slot of the returned _PBMDesign_ is cleared to prevent iterative application of the filtering critera. 

Finally, the `probeTrim` slot is an optional numeric vector of length 2. Again, this can be accessed and set using the `upbm::probeTrim` function.

```{r pbmdesign-show-probetrim}
probeTrim(pbm_8x60k_filtered)
```

Revisiting the design of the probes, notice that the probe sequences all end in a common 24nt primer sequence.

```{r, pbmdesign-show-seqs}
head(design(pbm_8x60k_filtered)$Sequence)
```

The `probeTrim` slot specifies the start and end positions in the sequences that should be kept to map K-mers to probes. By default, this is set to `(1, 36)` in the uPBM designs in _upbmAux_ to exclude the common 24nt primer sequence at the end of all probes. While the values may be modified to include some or all of the primer sequence, this has not been extensively evaluated. Probe trimming can be performed using `upbm::pbmTrimProbes`.

```{r pbmdesign-pbmtrimprobes}
pbm_8x60k_trimmed <- pbmTrimProbes(pbm_8x60k_filtered)
head(design(pbm_8x60k_trimmed))
```

## Modifying Objects

All of the slots described above can be modified or updated. For example, we can add an additional (redundant) rule to the `probeFilter` list of the example *PBMDesign*.

```{r pbmdesign-modify}
probeFilter(pbm_8x60k_v1) <- c(probeFilter(pbm_8x60k_v1),
                               Sequence = function(x) { x != "#N/A" })
```

We can also easily change the trimming boundaries.

```{r pbmdesign-modify-trim}
probeTrim(pbm_8x60k_filtered) <- c(1, 50)
pbm_8x60k_filtered
```

## New Objects

A new _PBMDesign_ object can be created by passing a data.frame of probe designs to the `upbm::PBMDesign` constructor. As stated above, at minimum, the data.frame must contain two columns, `"probeID"` and `"Sequence"`. 

```{r pbmdesign-new}
new_design <- data.frame(probeID = LETTERS[1:3],
                         Sequence = rep("AAA", 3))
pbmDesign <- PBMDesign(new_design)
pbmDesign
```

The `probeFilter` and `probeTrim` slots can also be specified directly to the `upbm::PBMDesign` constructor or modified after constructing the _PBMDesign_ object, as shown above.

# PBMExperiment

Probe-level PBM data is stored using the _PBMExperiment_ class, an extension of the [_SummarizedExperiment_](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html) class with additional slots to support working with _PBMDesign_ objects. 

An example _PBMExperiment_ object containing data from a series of HOXC9 experiments is included in the _upbmData_ package. We will only look at the _PBMExperiment_ containing Alexa488 scans. However, the probe-level data from Cy3 scans are similarly stored and included as a separate _PBMExperiment_ object in the package.

```{r pbme-load-example}
data(hoxc9alexa, package = "upbmData")
hoxc9alexa
```

A _SummarizedExperiment_ object is made up of three parts: **1.** assays (primary data, e.g. probe intensities), **2.** column metadata, and **3.** row metadata. Here, rows correspond to probes and columns correspond to scans. For more details on the _SummarizedExperiment_ class, see the corresponding package [vignette](https://bioconductor.org/packages/release/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html).

In addition to these three components, _PBMExperiment_ objects also include `probeFilter`, `probeTrim` and `probeCols` slots. The `probeFilter` and `probeTrim` slots of the _PBMExperiment_ class are identical to the slots described above as part of the _PBMDesign_ class. The final `probeCols` slot is a character vector specifying the columns in the row metadata corresponding to the `design` of a _PBMDesign_ object. Rather than storing the `design` as a separate slot in _PBMExperiment_ objects, the design is kept in the row metadata and the corresponding columns are recorded in `probeCols`.

As with _PBMDesign_ objects, these slots can be both accessed and modified. The `hoxc9alexa` object was constructed with the `pbm_8x60k_v1` probe design introduced above. We can see that the `probeTrim` and `probeFilter` values match what we saw above.

```{r pbme-access}
probeFilter(hoxc9alexa)
probeTrim(hoxc9alexa)
```

Additionally, we can see that the `probeCols` columns of the row metadata match the `design` slot of `pbm_8x60k_v1`.

```{r pbme-probecols}
probeCols(hoxc9alexa)
head(rowData(hoxc9alexa)[, probeCols(hoxc9alexa)])
```

Finally, the corresponding _PBMDesign_ object of a _PBMExperiment_ can be extracted by passing the _PBMExperiment_ object to `upbm::PBMDesign`.

```{r pbme-pbmdesign}
PBMDesign(hoxc9alexa)
```

We can check that this is identical to the _PBMDesign_ from above.

```{r, pbme-pbmdesign-identical}
identical(PBMDesign(hoxc9alexa), pbm_8x60k_v1)
```

The _PBMDesign_ associated values can be quickly modified or added using the associated setter function.

```{r pbem-set-design}
PBMDesign(hoxc9alexa) <- pbm_8x60k_v1
```

This can be useful when the design information needs to be added after the _PBMExperiment_ object has been created. However, this should be performed with caution as all prior design associated values will be overwritten. 

As with _PBMDesign_, calls to `upbm::pbmFilterProbes` and `upbm::pbmTrimProbes` can be used to filter and trim probes in the _PBMExperiment_ object.

```{r pbme-filter}
pbmFilterProbes(hoxc9alexa)
```

It is also important to note that whenever `broom::tidy` is called to convert assay data to a tidy format, `upbm::pbmFilterProbes` and `upbm::pbmTrimProbes` run unless `process = FALSE` is specified.

```{r pbme-tidy}
nrow(broom::tidy(hoxc9alexa))
nrow(broom::tidy(hoxc9alexa, process = FALSE))
```

## gpr2PBMExperiment

The easiest way to create a new _PBMExperiment_ object is by reading scan data from GPR files using `upbm::gpr2PBMExperiment`. Given a table of GPR file metadata, including paths to the files, the function will return the parsed GPR data as a _PBMExperiment_ object. 

In the case of the example dataset, the table used to read the GPR files is also included in the _upbmData_ package.

```{r}
data(hoxc9table, package = "upbmData")
head(hoxc9table, 5)
```

At a minimum, the full path to each GPR file **must** be specified in a column named `gpr`. The data.frame should also include any relevant metadata about the scan and sample (e.g. scan parameters or properties of the assayed protein). 

Both Alexa488 and Cy3 scans for the example dataset in the *upbmData* package are provided in `hoxc9table`. We will subset to just the Alexa488 scans.

```{r}
hoxc9table_alexa <- dplyr::filter(hoxc9table, type == "Alexa")
```

In addition to the sample table the corresponding _PBMDesign_ should also be specified when calling `upbm::gpr2PBMExperiment`. While not run here (since raw GPR files are not included with the package), the following call was made to generate the *PBMExperiment* object included in the _upbmData_ package.

```{r noeval-gpr2pbme, eval = FALSE}
hoxc9alexa <- gpr2PBMExperiment(scans = hoxc9table_alexa,
                                probes = pbm_8x60k_v1)
```

We can see that metadata included in the sample table are stored in the column metadata of the returned _PBMExperiment_ object.

```{r}
colData(hoxc9alexa)[1:5, ]
```

For _PBMExperiment_ objects created from GPR files, probe-level foreground intensities are stored in the `"fore"` assay. By default, (but optionally) a second assay, `"back"` (background) is also read in from the GPRs with the background intensities for each probe and sample. 

```{r}
assay(hoxc9alexa, "fore")[1:5, 1:5]
assay(hoxc9alexa, "back")[1:5, 1:5]
```

## New Objects

Alternatively, a new _PBMExperiment_ object can be constructed from existing _SummarizedExperiment_ and _PBMDesign_ objects. To demonstrate this, we can coerce `hoxc9alexa` to a _SummarizedExperiment_ object, stripping away all _PBMExperiment_-specific information.

```{r pbme-as-se}
se <- as(hoxc9alexa, "SummarizedExperiment")
se
```

Notice that the probe annotations are still kept in the row metadata. We can additionally remove this if we would like.

```{r se-drop-rowdata}
rowData(se) <- NULL
se
```

With the _SummarizedExperiment_ object, we can now create a new _PBMExperiment_ object using the `PBMExperiment(..)` constructor. As with `gpr2PBMExperiment(..)`, a _PBMDesign_ can be specified as an optional parameter.

```{r pbme-constructor}
PBMExperiment(se)

PBMExperiment(se, pbmDesign = pbm_8x60k_v1)
```

